Jane, John … Leslie? A Historical Method for Algorithmic Gender Detection

### 2014-09-12 article.Rmd
Implemented most of Cameron's suggestions

The problem arises from the fact that the link between gender and naming practices, like language itself, is not static. What about Jane and John Fay's sibling, Leslie? If Leslie was born in the past sixty years, chances are good that Leslie would be their sister. But if the three siblings were born in the early twentieth century, chances are good that Leslie would be their brother. That is because the gendered naming convention surrounding "Leslie" completely changed over the course of one hundred years. 92% of the babies born in the United States in 1900 who were named “Leslie” were male, while 96% of those identically named babies born in the year 2000 were female.

@@

Many other American names, such as Madison, Morgan, Sydney, and Kendall, have also changed gender over the past century. Such changes usually follow set patterns. In general, names tend to shift from being predominantly male to being predominantly female; the shift rarely moves in the other direction. In general, names that transition from male to female also became increasingly popular over time. For example, the name "Madison" changed from male to female suddenly after 1985. Whereas before it had been an infrequently used male name, after 1985 it skyrocketed in popularity, and by the early 2000s it was one of the ten most frequent names for baby girls. Broadly speaking, since the late twentieth century Americans girls are more likely than American boys to receive a "trendy" name, and once that name is associated with females it is unlikely to be given to male babies.^[The few exceptions of names which go from being female to male (such as "Jan") have a different causal mechanism. It seems likely that Jan as an English name is female but that Jan as a Scandinavian name is male. Because the name has remained relatively uncommon, other causes besides changing gender preferences can explain the change.]

@@

Researchers studying historical subjects, on the other hand, need to take into account changes in naming practices. If that same researcher wanted to compare the demographics of contemporary urban schools with those from the early twentieth century, contemporary datasets would misidentify Leslie Fay's gender. The "Leslie" problem is not just for researchers who think of themselves as historians. Anyone studying a timeframe longer than a few years, or anyone studying a group whose characteristics do not match the probably unknowable characteristics of the groups that comprise social media datasets will also encounter this problem. As of 2012, the average American has a lifespan of nearly seventy-nine years–--more than enough time for naming practices to change quite dramatically.^[CITATION] Predicting gender from first names therefore requires a fundamentally historical method.

Our solution to the "Leslie" problem is to create a software package that combines a predictive algorithm with a range of historical dataset suitable to various times and regions. The algorithm calculates the proportion of male or female uses of a name in a given birth year or range of years. It thus can provide not only a prediction of the gender of a name, but also an estimate of how likely that prediction is. For example 52 percent of the babies named Leslie born in the United States in 1950 were girls and 48% were boys. This allows researchers to conduct more nuanced analysis and to determine for themselves what proportion is acceptable for making a prediction of gender.

We recognize that gender is a fluid concept, both today and in the past, and that this method glosses over the complexities of historical definitions of gender and of the relationship between biological sex and gender. In some sense this is unavoidable since the large historical datasets which make this kind of analysis possible were created by state agencies. For the category of gender (as for race) states have offered a limited number of boxes to check, and those boxes extert a tremendous power to define state subjects.^[CITE One of the best-known theorists on gender is Judith Butler, *Gender Trouble* and Butler, *Bodies that Matter*; For the problems with the word "identity" in the humanities and social sciences, see Rogers Brubaker and Frederick Cooper, "Beyond 'Identity'"; Several historians have written about the American state and its drive to impose identities onto subjects: Margot Canady, The Straight State;  Peggy Pascoe, *What Comes Naturally*; Shane Landrum on birth certificates.] Nevertheless, this method is suitable for two reasons. First, researchers are likely to use this method in the aggregate rather than for individuals. The assumption that a name implies male or female (and that those are possible categories to be inferred) is more valid when applied to large groups of people than it does when applied to individuals. **I think this first line of reasoning is dangerous. This leaves us open to critiques of "big data" in general from humanists and theorists: that the process of aggregation is actually really dangerous and that assuming that individual identity and experience can be subsumed under the generalities of the crowd, etc. etc. I think the second reason is strong enough on its own.** Second, this method is neutral as to what the categories male or female actually mean in any given historical context. Researchers are free---indeed, obligated---to make that determination for themselves, and this package provides access to the underlying dataset so that researchers can interrogate every assumption that comes with them. This method does not argue against the complexities of historical understandings of gender, but as with any project it does require researchers who use the method to think through how the data was gathered, what it implies, and how it reflects social and cultural practices.

@@

Why do scholars need a historical method for inferring gender? Let's start with a comparison. One existing method for predicting gender is available in the [Natural Language Toolkit](http://www.nltk.org/) for Python.^[Steven Bird, Edward Loper, and Ewan Klein, *Natural Language Processing with Python* (O’Reilly, 2009), ch. 2. Chapter 6 of the NLTK book also describes a method of predicting gender from the last letter of the name, but this example is given for the sake of explaining supervised classification and not as an end in itself, so we do not have reason to address this a historical method.] The NLTK is an influential software package for scholarship because it provides an extraordinary range of tools for analyzing natural language. **I think we could develop this even more, if we have information about it. My impression is that it's the "industry standard" for a lot of folks, but it would be nice to give some evidence to then show why our intervention matters, and that the Kantrowitz method isn't just a straw man** Included in the NLTK are two lists of male and female names created by the computer scientist Mark Kantrowitz with assistance from Bill Ross.^[The [original lists of names](http://www.cs.cmu.edu/afs/cs/project/ai-repository/ai/areas/nlp/corpora/names/0.html) are available at the website of the computer science department at Carnegie Mellon University, and the NLTK implementation is described in its [documentation](http://www.nltk.org/howto/corpus.html).] These lists include `r kantrowitz_unique` unique names. Using the Kantrowitz names corpus in the NLTK, one could look up a name like Jane or John and find out that it is male or female. The Kantrowitz corpus provides the list of names.^[The examples below is drawn from our Gender package which implements the Kantrowitz dataset for comparsion's sake, though its use is not recommended for actual research. The Gender package combines the Kantrowitz corpus's two lists into a single list, identifying names that are ambiguous because they appear on both lists as `"either"`.]

@@

R and Python are the most commonly used languages for data analysis among digital humanists.^[**I completely agree, but do we have any evidence for this? Surveys or anything?**] Python's existing packages for gender prediction all implement a method similar to the NLTK or Genderize.io.^[Python library names] To our knowledge, [CRAN](http://cran.rstudio.com/) (the central repository for R packages) does not include any packages dedicated to gender prediction. In two of the most popular data analysis languages in the digital humanities there is currently no satisfactory existing method of gender prediction for historical and humanistic research.

To that end we have created the [Gender](https://github.com/ropensci/gender) package for R which includes both the predictive algorithm and an associated [genderdata](https://github.com/lmullen/gender-data-pkg) package containing various historical datasets. This R implementation is based on an [earlier Python implementation](https://github.com/cblevins/Gender-ID-By-Time) by [Cameron Blevins](http://www.cameronblevins.org/) and [Bridget Baird](http://www.conncoll.edu/directories/emeritus-faculty/bridget-baird/). The Gender package is affiliated with [rOpenSci](http://ropensci.org/), an initiative that supports reproducible research and open data for scientists using R. The rOpenSci advisory board has provided code review and publicity for the project.

@@

The Gender package currently includes two datasets which make it suitable for studying the United States from the first federal census in 1790 onwards.^[The authors of the package are both American historians, which explains why the initial release of the package focuses on that country. We have plans, however, to extend the underlying data to six other European and North American countries included in the [North Atlantic Population Project](https://www.nappdata.org/napp/) for the nineteenth century. We also welcome contributions of datasets for other regions and times from other scholars.] The first dataset contains names of applicants for Social Security and available from Data.gov. The second dataset contains names gathered in the decennial censuses and available from the IPUMS-USA (Integrated Public Use Microdata Series) service from the Minnesota Population Center at the University of Minnesota.^[**citation?**]

The Social Security Administration "[Baby Names](http://catalog.data.gov/dataset/baby-names-from-social-security-card-applications-national-level-data)" dataset was created as a result of the Social Security Act of 1935 during the New Deal. ^[Social Security Administration, "[Baby Names from Social Security Card Applications-National Level Data](http://catalog.data.gov/dataset/baby-names-from-social-security-card-applications-national-level-data)," Data.gov dataset ID US-GOV-SSA-338 (accessed August 31, 2014). The Social Security Administration also provides a related dataset which contains the names of applicants broken down by state and year since 1910. Social Security Administration, "[Baby Names from Social Security Card Applications-Data by State and District of Columbia](http://catalog.data.gov/dataset/baby-names-from-social-security-card-applications-data-by-state-and-district-of-)," Data.gov dataset ID US-GOV-SSA-339 (accessed August 31, 2014).] This dataset contains a list of first names with the number of male and female uses of the name in every year beginning with 1880 and ending with 2012. The SSA list includes any name which was used more than five times in a given year, thereby capturing all but the most infrequently used names from each year. The description "baby names" provided by the SSA is a serious misnomer. When Social Security first became available during the New Deal, its first beneficiaries were adults past or near retirment age. The dataset goes back to 1880, the birth year for a 55 year-old adult when Social Security was enacted. **Okay, so that means we need to tweak some stuff where we use the word "baby" earlier, yes?** Even after 1935, registration at birth for Social Security was not mandatory until 1986. As we will demonstrate below, the way in which the data was gathered requires an adjustment to our predictions of gender.^[CITE history of SSA. On the history of birth registration see Shane Landrum, "The State's Big Family Bible: Birth Certificates, Personal Identity, and Citizenship in the United States, 1840--1950" (PhD dissertation, Brandeis University, 2014). It is well-known among American historians that Social Security was written to exclude farmworkers as a means of also excluding African Americans. In examining the dataset we have not found any reason to believe that this policy noticably skewed the prediction of gender from names.]

@@

Our analysis focuses on one of the bedrocks of the historical profession: scholarly research. We begin with the history dissertation, often the defining scholarly output of a historian's early career. The completion of a dissertation marks a standardized moment of transition out of the "training" phase of the historical profession. Data supplied by ProQuest provided roughly [80,000] PhD(?) dissertations completed since [1950]. Identifying the gender of their authors gives a sense for the number of women and men who are completing PhD-level training in history each year. Our program uses what year a historian wrote their dissertation to estimate a period for when they might have been born: a historian who completed their dissertation in 1980 was likely born between ____ and _____. Using this temporal information, we are better able to infer their gender and chart how the larger representation of women and men changes year-by-year.

@@

Another way to examine this trend is to look at the proportion of male and female dissertation authors over time, which smoothes out changes in the absolute number of dissertations produced. The proportion of dissertations written by women has steadily increased over the past half-century, a change that began in the late 1960s and continued through the early 2000’s. Since that point, the proportion of dissertations written by women has largely plateaued at a few percentage points below the proportion written by men. Female historians have achieved something approaching parity with male historians in terms of how many women and women complete dissertations each year.

>>>>

The problem arises from the fact that the link between gender and naming practices, like language itself, is not static. What about Jane and John Fay's sibling, Leslie? If Leslie was born in the past sixty years, chances are good that Leslie would be their sister. But if the three siblings were born in the early twentieth century, chances are good that Leslie would be their brother. That is because the gendered naming convention surrounding "Leslie" completely changed over the course of one hundred years. 92% of the babies born in the United States in 1900 who were named "Leslie" were male, while 96% of those identically named babies born in the year 2000 were female.

@@

Many other names as used in the United States, such as Madison, Morgan, Sydney, and Kendall, have also changed gender over the past century. Such changes usually follow set patterns. In general, names tend to shift from being predominantly male to being predominantly female; the shift rarely moves in the other direction. In general, names that transition from male to female also became increasingly popular over time. For example, the name "Madison" changed from male to female suddenly after 1985. Whereas before it had been an infrequently used male name, after 1985 it skyrocketed in popularity, and by the early 2000s it was one of the ten most frequently used names for baby girls. Broadly speaking, since the late twentieth century Americans girls are more likely than American boys to receive a "trendy" name, and once that name is associated with females it is unlikely to be given to male babies.^[The few exceptions of names which go from being female to male (such as "Jan") have a different causal mechanism. It seems likely that Jan as an English name is female but that Jan as a Scandinavian name is male. Because the name has remained relatively uncommon, other causes besides changing gender preferences can explain the change.]

@@

Researchers studying historical subjects, on the other hand, need to take into account changes in naming practices. If that same researcher wanted to compare the demographics of contemporary urban schools with those from the early twentieth century, contemporary datasets would misidentify Leslie Fay's gender. The "Leslie" problem is not just for researchers who think of themselves as historians. Anyone studying a period longer than a few years, or anyone studying a group whose characteristics do not match the probably unknowable characteristics of the groups that comprise social media datasets will also encounter this problem. As of 2012, the average American has a lifespan of nearly seventy-nine years–--more than enough time for naming practices to change quite dramatically.^[World Bank, *World Bank Open Data*, "Life expectancy at birth, total (years)," <http://data.worldbank.org/indicator/SP.DYN.LE00.IN> (accessed September 12, 2014).] Predicting gender from first names therefore requires a fundamentally historical method.

Our solution to the "Leslie" problem is to create a software package that combines a predictive algorithm with a several historical datasets suitable to various times and regions. The algorithm calculates the proportion of male or female uses of a name in a given birth year or range of years. It thus can provide not only a prediction of the gender of a name, but also an estimate of how likely that prediction is. For example 52 percent of the babies named Leslie born in the United States in 1950 were girls and 48 percent were boys. This allows researchers to conduct more nuanced analysis and to determine for themselves what proportion is acceptable for making a prediction of gender.

We recognize that gender is a fluid concept, both today and in the past, and that this method glosses over the complexities of historical definitions of gender and of the relationship between biological sex and gender. In some sense this is unavoidable since the large historical datasets which make this kind of analysis possible were created by state agencies. For the category of gender (as for race) states have offered a limited number of boxes to check, and those boxes extert a tremendous power to define state subjects.^[One of the best-known theorists on gender is Judith Butler, *Gender Trouble* and Butler, *Bodies that Matter*; For the problems with the word "identity" in the humanities and social sciences, see Rogers Brubaker and Frederick Cooper, "Beyond 'Identity'"; Several historians have written about the American state and its drive to impose identities onto subjects: Margot Canady, The Straight State;  Peggy Pascoe, *What Comes Naturally*; Shane Landrum on birth certificates.] Nevertheless, this method is neutral as to what the categories male or female actually mean in any given historical context. Researchers are free---indeed, obligated---to make that determination for themselves, and this package provides access to the underlying dataset so that researchers can interrogate every assumption that comes with them. This method does not argue against the complexities of historical understandings of gender, but as with any project it does require researchers who use the method to think through how the data was gathered, what it implies, and how it reflects social and cultural practices.

@@

Why do scholars need a historical method for inferring gender? Let's start with a comparison. One existing method for predicting gender is available in the [Natural Language Toolkit](http://www.nltk.org/) for Python.^[Steven Bird, Edward Loper, and Ewan Klein, *Natural Language Processing with Python* (O’Reilly, 2009), ch. 2. Chapter 6 of the NLTK book also describes a method of predicting gender from the last letter of the name, but this example is given for the sake of explaining supervised classification and not as an end in itself, so we do not have reason to address this a historical method.] The NLTK is an influential software package for scholarship because it provides an extraordinary range of tools for analyzing natural language.^[The NLTK book (the most common way to cite the software) has been cited 812 times according to [Google Scholar](http://scholar.google.com/scholar?cites=3735380965977150091&as_sdt=5,47&sciodt=0,47&hl=en). Because in the humanities (unlike in other fields) the practice of citing even software which determines the outcome of an analysis is regrettably rare, even nonexistent, such metrics are difficult to determine. Yet searching of journals such as *Digital Humanities Quarterly* and the *Journal of Digital Humanities* reveals that natural language processing, for which the NLTK is the foremost implementation, is one of the key methodological categories of the digital humanities.] Included in the NLTK are two lists of male and female names created by the computer scientist Mark Kantrowitz with assistance from Bill Ross.^[The [original lists of names](http://www.cs.cmu.edu/afs/cs/project/ai-repository/ai/areas/nlp/corpora/names/0.html) are available at the website of the computer science department at Carnegie Mellon University, and the NLTK implementation is described in its [documentation](http://www.nltk.org/howto/corpus.html).] These lists include `r kantrowitz_unique` unique names. Using the Kantrowitz names corpus in the NLTK, one could look up a name like Jane or John and find out that it is male or female. The Kantrowitz corpus provides the list of names.^[The examples below is drawn from our Gender package which implements the Kantrowitz dataset for comparsion's sake, though its use is not recommended for actual research. The Gender package combines the Kantrowitz corpus's two lists into a single list, identifying names that are ambiguous because they appear on both lists as `"either"`.]

@@

R and Python are the most commonly used languages for data analysis among digital humanists.^[Surveys and analysis of useage point out the growth of R and the continuing popularity of Python among "data science" generally. See, for example, Robert A. Muenchen, "The Popularity of Data Analysis Software," *r4stats.com*, <http://r4stats.com/articles/popularity/> (accessed September 12, 2014); Gregory Piatetsky, "Top Languages for analytics, data mining, data science," *KDN Nuggets*, <http://www.kdnuggets.com/2013/08/languages-for-analytics-data-mining-data-science.html> (accessed September 12, 2014).] Python's existing packages for gender prediction all implement a method similar to the NLTK or Genderize.io.^[Python library names] To our knowledge, [CRAN](http://cran.rstudio.com/) (the central repository for R packages) does not include any packages dedicated to gender prediction. In two of the most popular data analysis languages in the digital humanities there is currently no satisfactory existing method of gender prediction for historical and humanistic research.

To that end we have created the [Gender](https://github.com/ropensci/gender) package for R which includes both the predictive algorithm and an associated [genderdata](https://github.com/lmullen/gender-data-pkg) package containing various historical datasets. This R implementation is based on an [earlier Python implementation](https://github.com/cblevins/Gender-ID-By-Time) by [Cameron Blevins](http://www.cameronblevins.org/) and [Bridget Baird](http://www.conncoll.edu/directories/emeritus-faculty/bridget-baird/). The Gender package is affiliated with [rOpenSci](http://ropensci.org/), an initiative that supports reproducible research and open data for scientists using R. The rOpenSci advisory board has provided code review and publicity for the project.^[The gender package depends especially on R Core Team, *R: A language and environment for statistical computing*. R Foundation for Statistical Computing, Vienna, Austria (2014). <http://www.R-project.org/>;  Hadley Wickham and Romain Francois, *dplyr: dplyr: a grammar of data manipulation*. R package version 0.2 (2014): <http://CRAN.R-project.org/package=dplyr>. The sources of the data are cited below.]

@@

The Gender package currently includes two datasets which make it suitable for studying the United States from the first federal census in 1790 onwards.^[The authors of the package are both American historians, which explains why the initial release of the package focuses on that country. We have plans, however, to extend the underlying data to six other European and North American countries included in the [North Atlantic Population Project](https://www.nappdata.org/napp/) for the nineteenth century. We also welcome contributions of datasets for other regions and times from other scholars.] The first dataset contains names of applicants for Social Security and available from Data.gov. The second dataset contains names gathered in the decennial censuses and available from the IPUMS-USA (Integrated Public Use Microdata Series) service from the Minnesota Population Center at the University of Minnesota.

The Social Security Administration "[Baby Names](http://catalog.data.gov/dataset/baby-names-from-social-security-card-applications-national-level-data)" dataset was created as a result of the Social Security Act of 1935 during the New Deal. ^[Social Security Administration, "[Baby Names from Social Security Card Applications-National Level Data](http://catalog.data.gov/dataset/baby-names-from-social-security-card-applications-national-level-data)," Data.gov dataset ID US-GOV-SSA-338 (accessed August 31, 2014). The Social Security Administration also provides a related dataset which contains the names of applicants broken down by state and year since 1910. Social Security Administration, "[Baby Names from Social Security Card Applications-Data by State and District of Columbia](http://catalog.data.gov/dataset/baby-names-from-social-security-card-applications-data-by-state-and-district-of-)," Data.gov dataset ID US-GOV-SSA-339 (accessed August 31, 2014).] This dataset contains a list of first names with the number of male and female uses of the name in every year beginning with 1880 and ending with 2012. The SSA list includes any name which was used more than five times in a given year, thereby capturing all but the most infrequently used names from each year. The description "baby names" provided by the SSA is a serious misnomer. When Social Security first became available during the New Deal, its first beneficiaries were adults past or near retirment age. The dataset goes back to 1880, the birth year for a 55 year-old adult when Social Security was enacted. Even after 1935, registration at birth for Social Security was not mandatory until 1986. As we will demonstrate below, the way in which the data was gathered requires an adjustment to our predictions of gender.^[CITE history of SSA. On the history of birth registration see Shane Landrum, "The State's Big Family Bible: Birth Certificates, Personal Identity, and Citizenship in the United States, 1840--1950" (PhD dissertation, Brandeis University, 2014). It is well-known among American historians that Social Security was written to exclude farmworkers as a means of also excluding African Americans. In examining the dataset we have not found any reason to believe that this policy noticably skewed the prediction of gender from names.]

@@

Our analysis focuses on one of the bedrocks of the historical profession: scholarly research. We begin with the history dissertation, often the defining scholarly output of a historian's early career. The completion of a dissertation marks a standardized moment of transition out of the "training" phase of the historical profession. Data supplied by ProQuest provided roughly [80,000] PhD(?) dissertations completed since [1950].^[ProQuest's dataset does not permit us to limit dissertations just to dissertations produced in history departments. These dissertations include dissertations *about* history. For a justification of this approach, see a [series of blog posts](http://lincolnmullen.com/research/history-dissertations/) exploring the dataset. The American Historical Association's dataset of dissertations relies on reporting from departments and does not have the chronological range of the ProQuest data. Nevertheless, applying the same method to the AHA's dataset does not provide any reason for thinking that the proportion authorship by gender over time is substantially different when limited to just history department dissertations. American Historical Association, *Directory of History Dissertations*, https://secure.historians.org/pubs/dissertations/. Data export provided to author by Liz Townsend, March 3, 2014.] Identifying the gender of their authors gives a sense for the number of women and men who are completing PhD-level training in history each year. Our program uses the year a historian wrote his or her dissertation to estimate a period for when they might have been born: a historian who completed their dissertation in 1980 was likely born between ____ and _____. Using this temporal information, we are better able to infer their gender and chart how the larger representation of women and men changes year-by-year.

@@

Another way to examine this trend is to look at the proportion of male and female dissertation authors over time, which smoothes out changes in the absolute number of dissertations produced. The proportion of dissertations written by women has steadily increased over the past half-century, a change that began in the late 1960s and continued through the early 2000's. Since that point, the proportion of dissertations written by women has largely plateaued at a few percentage points below the proportion written by men. Female historians have achieved something approaching parity with male historians in terms of how many women and women complete dissertations each year.
